---
title: An R Markdown document converted from "1. GTFS RT aggregation.ipynb"
output: html_document
---

# GTFS RT aggregation

```{r}
# Native R implementation of the GTFS RT aggregation workflow
library(jsonlite)
library(lubridate)
library(geosphere)
library(dplyr)
library(readr)
library(sf)
```

## Parameters

**Important!** Before running, make sure to have `stop_times.txt` GTFS file at `stop_times` folder, with both `.txt` and `.csv` extensions!

```{r}
folder_path <- "../data/cm_20260413_220260430_business/updates"
output_folder <- "../data/cm_20260413_220260430_business"
stop_times <- "../data/cm_20260413_220260430_business/stop_times.txt"

SPEED_STOPPED <- 6 # Only consider updates with <6km/h speed (100 meters difference in 1 min)
THRESHOLD_UPDATES_PER_TRIP_MIN = 20
THRESHOLD_SPEED_MAX = 140
THRESHOLD_TIME_BETWEEN_UPDATES_MAX = 60
METRIC_CRS <- 3763 # Portugal
timezone <- lubridate::tz(ymd_hms("2026-01-01 00:00:00", tz = "Europe/Lisbon"))
```

## Methods

```{r}
process_json <- function(data, filename, RECORDS_LIST, agency_id = NULL) {
  day <- strsplit(filename, "_", fixed = TRUE)[[1]][1]

  for (e in data$entities) {
    vehicle <- e$vehicle
    if (is.null(vehicle$trip)) {
      message("Skipping vehicle with no trip information in file: ", filename)
      next
    }

    trip_id <- vehicle$trip$trip_id
    route_id <- vehicle$trip$route_id
    trip_id_day <- paste0(trip_id, "_", day)

    

    # If previous has same timestamp, skip (to avoid duplicates)
    if (is.null(RECORDS_LIST[[trip_id_day]])) {
      RECORDS_LIST[[trip_id_day]] <- data.frame()
    }
    if (nrow(RECORDS_LIST[[trip_id_day]]) > 0 && RECORDS_LIST[[trip_id_day]] |>
        slice(1) |>
        pull(timestamp) == vehicle$timestamp) 
    {
      #message("Skipping duplicate timestamp for trip_id: ", trip_id, " on day: ", day)
      next
    }

    RECORDS_LIST[[trip_id_day]] <- bind_rows(RECORDS_LIST[[trip_id_day]], data.frame(
      trip_id = trip_id,
      route_id = route_id,
      day = day,
      latitude = vehicle$position$latitude,
      longitude = vehicle$position$longitude,
      timestamp = vehicle$timestamp,
      current_stop_sequence = vehicle$current_stop_sequence
    ))
  }

  RECORDS_LIST
}
```

```{r}
process_stop_times <- function(file_path) {
  df <- readr::read_csv(file_path, show_col_types = FALSE)
  df$stop_sequence <- as.integer(df$stop_sequence)
  df %>%
    group_by(trip_id) %>%
    summarise(min = min(stop_sequence, na.rm = TRUE), max = max(stop_sequence, na.rm = TRUE), .groups = "drop")
}
```

```{r}
get_trip_stops <- function(trips_stops, trip_id) {
  filtered <- trips_stops[trips_stops$trip_id == trip_id, ]

  if (nrow(filtered) > 0) {
    c(as.integer(filtered$min[[1]]), as.integer(filtered$max[[1]]))
  } else {
    message("Trip ID not found: ", trip_id)
    c(NA_integer_, NA_integer_)
  }
}
```

```{r}
process_trip <- function(RECORDS, TRIPS_STOPS) {
  trips_processed <- RECORDS |>
    filter(trip_id %in% TRIPS_STOPS$trip_id) |> # Only valid trip_id
    sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE) |>
    sf::st_transform(crs = METRIC_CRS) |>
    group_split(trip_id) |>
    purrr::map_dfr(function(trip_df) {
      trip_trips_stops = TRIPS_STOPS |> filter(trip_id == trip_df$trip_id[1])
      return(trip_df |> mutate(
        time_since_prev_sec = timestamp - lag(timestamp),
        euclidean_distance_since_prev_meters = units::drop_units(st_distance(lag(geometry), geometry, by_element = TRUE)),
        euclidean_speed_kmh = (euclidean_distance_since_prev_meters / 1000) / (time_since_prev_sec / 3600),
        manhattan_distance_since_prev_meters = distHaversine(cbind(lag(longitude), lag(latitude)), cbind(longitude, latitude)),
        manhattan_speed_kmh = (manhattan_distance_since_prev_meters / 1000) / (time_since_prev_sec / 3600),
      ) |> filter(
        !is.na(euclidean_speed_kmh)
        & # Remove first and last stop (as bus tends to stop there for a while)
          current_stop_sequence > trip_trips_stops$min[[1]] &
          current_stop_sequence < trip_trips_stops$max[[1]]
      ) |> select(-geometry))
    })
}
```

## Analysis

### 0. Load stop_times and for each, get min and max stop_sequence

```{r}
trips_stops <- process_stop_times(stop_times)
print(trips_stops)
# Save to a CSV file if needed
# readr::write_csv(trips_stops, "trip_stop_sequences.csv")
```

```{r}
mi_ma <- get_trip_stops(trips_stops, "6348_20260101_106_0_8")
print(mi_ma[1])
print(mi_ma[2])

mi_ma <- get_trip_stops(trips_stops, "abc")
print(mi_ma[1])
print(mi_ma[2])
```

### 1. Aggregate JSONs (one per minute) by trip_id, with record of all updates per trip

```{r}
print("\n1. Starting aggregation of GTSF-RT files...")
if (!dir.exists(folder_path)) {
  stop(sprintf("Folder '%s' does not exist.", folder_path))
}

json_files <- list.files(folder_path, pattern = "\\.json$", full.names = TRUE)
json_files_dates <- sort(unique(substr(basename(json_files), 1, 8)))
for (date in json_files_dates) {
  RECORDS_LIST <- list()
  message(sprintf("Processing date: %s", date))
  date_files <- json_files[substr(basename(json_files), 1, 8) == date]
  for (file_path in date_files) {
    filename <- basename(file_path)
    tryCatch({
      data <- jsonlite::fromJSON(file_path, simplifyVector = FALSE)
      RECORDS_LIST <- process_json(data, filename, RECORDS_LIST)
    }, error = function(e) {
      message(sprintf("Unexpected error processing file %s: %s", filename, e$message))
    })
  }
  
  RECORDS <- data.frame()
  # Combine all individual trip data frames into a single data frame
  RECORDS <- do.call(rbind, RECORDS_LIST)
  
  avg_updates <- RECORDS |> group_by(trip_id) |>
    summarise(updates_count = n(), .groups = "drop") |>
    summarise(avg_updates = mean(updates_count)) |>
    pull(avg_updates)
  print(sprintf("\nDONE! Processed %d updates, corresponding to %d individual trips, with an average number of %s updates per trip\n", nrow(RECORDS), length(unique(RECORDS$trip_id)), avg_updates))
  write.csv(RECORDS, file.path(output_folder, sprintf("updates_raw_%s.csv", date)), row.names = FALSE)

  # Remove variables and clear memory
  rm(RECORDS_LIST, RECORDS, avg_updates)
  gc()  # Call garbage collector to free up memory
}
```

### 2. Compute speeds for each update (relating each update with previous, Euclidean distance)

```{r}
print("\n2. Starting computation of average speed between updates...")
for (date in json_files_dates) {
    RECORDS <- read.csv(file.path(output_folder, sprintf("updates_raw_%s.csv", date)))
    result <- process_trip(RECORDS, trips_stops) |> 
      st_drop_geometry()
    
    metrics = result |> group_by(trip_id) |>
      summarise(updates_count = n(), .groups = "drop") |>
      filter(updates_count > THRESHOLD_UPDATES_PER_TRIP_MIN)
    summary(metrics$updates_count)
    
    result <- result |>
      # Consider only trips with +20 updates
      filter(trip_id %in% metrics$trip_id) |>
      # Consider only updates with <60 sec between them
      filter(time_since_prev_sec < THRESHOLD_TIME_BETWEEN_UPDATES_MAX) |>
      # Filter outlier speeds 
      filter(euclidean_speed_kmh < THRESHOLD_SPEED_MAX)
    
    message(sprintf("DONE! Processed speed for date: %s, with average speed of %.2f km/h", date, mean(result$euclidean_speed_kmh, na.rm = TRUE)))

    write.csv(result |> sf::st_drop_geometry(), file.path(output_folder, sprintf("updates_with_speed_%s.csv", date)), row.names = FALSE)
}
```
