---
title: "GTFS Real-Time Shape Distance Analysis"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

This analysis evaluates the distance between real-time vehicle positions and the closest point on their assigned geometry shape. This metric helps understand positioning accuracy and potential GPS/network issues.

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(knitr)
```

```{r}
rt_collection_files = I(list(as.character(list.files(
    "../../data/cm_20260413_220260430_business/processing_speed_shape_distance_osm_thresholds_circular_fix",
    pattern = "^updates_with_speed_.*\\.csv$",
    full.names = TRUE
))))

# rt_collection_files = I(list("../../data/cm_20260413_220260430_business/processing_speed_shape_distance_osm_thresholds_circular_fix/updates_with_speed_20260413.csv"))

rt_files <- as.character(rt_collection_files[[1]])
rt_collection_raw <- dplyr::bind_rows(lapply(rt_files, read.csv))
```



## Global Metrics

Overview of distance dispersion across all real-time position records.

```{r}
# Summary statistics for distance_to_closest_on_geometry
summary_stats <- rt_collection_raw |>
  summarise(
    n_records = n(),
    mean_distance = mean(distance_to_closest_on_geometry, na.rm = TRUE),
    median_distance = median(distance_to_closest_on_geometry, na.rm = TRUE),
    sd_distance = sd(distance_to_closest_on_geometry, na.rm = TRUE),
    min_distance = min(distance_to_closest_on_geometry, na.rm = TRUE),
    max_distance = max(distance_to_closest_on_geometry, na.rm = TRUE),
    p25 = quantile(distance_to_closest_on_geometry, 0.25, na.rm = TRUE),
    p75 = quantile(distance_to_closest_on_geometry, 0.75, na.rm = TRUE),
    p95 = quantile(distance_to_closest_on_geometry, 0.95, na.rm = TRUE),
    p99 = quantile(distance_to_closest_on_geometry, 0.99, na.rm = TRUE),
    na_count = sum(is.na(distance_to_closest_on_geometry))
  )

knitr::kable(
  summary_stats,
  caption = "Global summary statistics for distance to closest shape point"
)
```



## Distance Thresholds Analysis

Percentage of records within acceptable distance thresholds.

```{r}
# Threshold analysis
thresholds <- c(5, 10, 25, 50, 100)

threshold_analysis <- data.frame(
  threshold_meters = thresholds,
  count_within = sapply(thresholds, function(t) 
    sum(rt_collection_raw$distance_to_closest_on_geometry <= t, na.rm = TRUE)),
  percentage = sapply(thresholds, function(t) 
    round(100 * sum(rt_collection_raw$distance_to_closest_on_geometry <= t, na.rm = TRUE) / 
          nrow(rt_collection_raw), 2))
)

knitr::kable(
  threshold_analysis,
  caption = "Threshold analysis: records within distance limits"
)
```



## Analysis by Day and Trip

Detailed breakdown of distance metrics grouped by day and trip_id to identify patterns and problematic routes.

```{r}
# Prepare data with day extraction
rt_data_daily <- rt_collection_raw |>
  filter(!is.na(trip_id) & !is.na(distance_to_closest_on_geometry))

# Summary statistics by day and trip_id
day_trip_stats <- rt_data_daily |>
  group_by(day, trip_id) |>
  summarise(
    n_records = n(),
    mean_distance = mean(distance_to_closest_on_geometry, na.rm = TRUE),
    median_distance = median(distance_to_closest_on_geometry, na.rm = TRUE),
    sd_distance = sd(distance_to_closest_on_geometry, na.rm = TRUE),
    max_distance = max(distance_to_closest_on_geometry, na.rm = TRUE),
    p95_distance = quantile(distance_to_closest_on_geometry, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_distance))

# Show top 10 worst performing trip/day combinations
knitr::kable(
  head(day_trip_stats, 10),
  caption = "Top 10 trip/day combinations with highest mean distance to shape"
)

# Daily aggregates
daily_summary <- day_trip_stats |>
  group_by(day) |>
  summarise(
    n_trips = n_distinct(trip_id),
    n_records = sum(n_records),
    mean_distance = mean(mean_distance, na.rm = TRUE),
    median_distance = median(median_distance, na.rm = TRUE),
    max_distance = max(max_distance, na.rm = TRUE),
    p95_distance = mean(p95_distance, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(day)

knitr::kable(
  daily_summary,
  caption = "Daily summary across trips"
)
```


